Medical Concept Extraction

### 2012-12-17 paper/paper.tex
Final pass through N-gram Features.

\subsection{Sentence Features}
Even though the primary goal of our system is to tag concepts at the word level, these individual words often need to be considered in context. Before extracting the word-level features described in 5.2, we first analyze medical records at the sentence level and assign features to our words based on their usage within those sentences.

\subsubsection{Part of Speech}
A fruitful strategy for any concept extractor is to first tag words with labels indicating their Part of Speech in context. In our case, for example, whether an ambiguous term is used as an adjective versus a noun could distinguish it as part of a test rather than a problem. We used the NTLK max entropy Part of Speech Tagger, which is trained on the Penn TreeBank to tag individual sentences. We then iterate over the returned tags and add these as word-level features, adding another dimension to our feature set for each Part of Speech tag.

\subsubsection{WordNet Lemmatizer}
Unlike the Porter Stemmer and the Lancaster Stemmer, which consider words independent of context, the WordNet Lemmatizer requires words to be processed at the sentence level in order to extract their root form. The WordNet Lemmatizer uses the part of speech tags for individual words, which are assign via NLTK as previously mentioned, in conjunction with a subset of tags used for morphology to extract stems from each word. WordNet references a database of word stems, and uses two types of processes to try to convert input strings into a form present in the database: A list of inflectional endings based on the word's syntactic tag is referenced in search of a match that can be detached from the word. An exception list is also referenced in search for an inflected form. WordNet first checks for exceptions and then uses rules for detachment to transform the word--after each transformation the result is referenced against the stem database until a match is eventually found.

\subsubsection{Formatted Text}
When reading through our training data, we noticed that test results are written in a variety of distinct forms that clearly separate these terms from problem or treatment terms. Since test result forms often spent multiple words such was "O2 sat -- 98\%" or "screen was positive". We wrote a series of regular expressions to capture these patterns with the test term as the first term in the sequence. We iterate over each word in the sentence and run our RegEx suite on the inclusive righthand sequence starting from that word. This feature adds only one additional dimension, and improved our CRF F1 test score from .85 to .90 by boosting our test term detection appreciably.

\subsection{N-gram Features}
Although our primary goal is to classify $individual$ words, a single word in a medical record entry is hardly independent from the next.  More specifically, medical concepts can often span more than one word--we may not be interested in detecting these exact spans, but features from certain words provide insight into how to label their neighbors. For example, ambiguous words such as "heart" could precede "rate" or "disease", and making that distinction between test and problem is much easier when we examine the subsequent word in the sentence containing "heart". A neighboring word can serve as an indicator even if it is not part of the same concept span. For example, words like "showed" and its synonyms are non-medical, but tend to follow a test term. Likewise, "rule out" and "history of" tend to precede problems. With this in mind, we observe all features of immediately neighboring words.

\subsubsection{Previous}
We take a basic approach to context features by considering the immediately preceding word for each individual word. Given a word with the feature set previously described, we simply duplicate each feature with the corresponding value for it's previous neighbor and rename these features $prev\_$<NAME> (e.g. $prev\_$length).  This serves to double the size of our feature space.
\subsubsection{Next}
Likewise, we duplicate the subsequent word features, which are relabeled as $next\_$<NAME>. At this point, our feature space is tripled, but was already quite large--hence, we decided only to consider surrounding words at a bigram level. Nevertheless, this relatively straightforward step brought out CRF F1 testing score up to 0.98.

>>>>

\subsection{Sentence Features}
Even though the primary goal of our system is to tag concepts at the word level, these individual words often need to be considered in context. After extracting the word-level features described in 5.2, we then analyze medical records at the sentence level and assign features to our words based on their usage within those sentences.

\subsubsection{Part of Speech}
A fruitful strategy for any concept extractor is to first tag words with labels indicating their part of speech. In our case, for example, whether an ambiguous term is used as an adjective versus a noun could distinguish it as part of a test rather than a problem. We used the NTLK max entropy Part of Speech Tagger, which is trained on the Penn TreeBank to tag each sentences. We then iterate over the returned tags and add these as word-level features, adding several dimensions to our feature set, one for each part of speech tag.

\subsubsection{WordNet Lemmatizer}
Unlike the Porter Stemmer and the Lancaster Stemmer, which consider words independent of context, the WordNet Lemmatizer requires words to be processed at the sentence level in order to extract their root form. The WordNet Lemmatizer uses the part of speech tags for individual words, which are assign via NLTK as previously mentioned, in conjunction with a subset of tags used for morphology to extract stems from each word. WordNet references a database of word stems, and uses two types of processes to try to convert input strings into a form present in the database: A list of inflectional endings based on the word's syntactic tag is referenced in search of a match that can be detached from the word. An exception list is also referenced in search for an inflected form. WordNet first checks for exceptions and then uses rules for detachment to transform the word. After each transformation the result is referenced against the stem database until a match is eventually found.

\subsubsection{Formatted Text}
When reading through our training data, we noticed that test results are written in a variety of distinct forms that clearly separate these terms from problem or treatment terms. Since test result forms often spanned multiple words such was ``O2 sat -- 98\%'' or ``screen was positive''. We wrote a series of regular expressions to capture these patterns with the test term as the first term in the sequence. We iterate over each word in the sentence and run our RegEx suite on the inclusive righthand sequence starting from that word. This feature adds only one additional dimension, and improved our toy test CRF F1 test score from .85 to .90 by boosting our test term detection appreciably.

\subsection{N-gram Features}
Again, although our primary goal is to classify individual words, a single word in a medical record entry is hardly independent from the next.  More specifically, medical concepts can often span more than one word. We may not be interested in detecting these exact spans, but features from certain words provide insight into how to label their neighbors. For example, ambiguous words such as "heart" could precede "rate" or "disease", and making that distinction between test and problem is much easier when we examine the subsequent word in the sentence containing "heart". A neighboring word can serve as an indicator even if it is not part of the same concept span. For example, words like "showed" and its synonyms are non-medical, but tend to follow a test term. Likewise, "rule out" and "history of" tend to precede problems. With this in mind, we observe all features of immediately neighboring words.

\subsubsection{Previous}
We take a basic approach to context features by considering the immediately preceding word for each individual word. Given a word with the feature set previously described, we simply duplicate each feature with the corresponding value for it's previous neighbor and rename these features $prev\_$<NAME> (e.g. $prev\_$length).  This serves to double the size of our feature space.

\subsubsection{Next}
Likewise, we duplicate the subsequent word features, which are relabeled as $next\_$<NAME>. At this point, our feature space is tripled, but was already quite large--hence, we decided only to consider surrounding words at in this ``bigram''-like context. Nevertheless, this relatively straightforward step brought our toy test CRF F1 testing score up to 0.98.
